Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
wordsmithing, a few additions, and a steaming dump of an abstract. We need to lose half a page.

\begin{abstract}
HELLO?  DRAFT OF AN ABSTRACT?
\end{abstract}

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a query sequence is called 
\textit{homology detection}.

@@

Homology-detection software most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. 
This alignment represents structural
correspondence among individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.

An alignment may contain \emph{gaps} (notated as dashes in
\figref{alignment}) for some proteins.
A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
protein~2 lost its amino acid in position~$j$, or 
other proteins gained an amino acid in position~$j$.
If~column~$j$ contains few gaps, 
it~is considered a \emph{consensus column},
and the few proteins with gaps probably lost amino acids via
\emph{deletions}. 
If~column~$j$ contains \emph{mostly} gaps, 
it~is considered a \emph{non-consensus column},
and the few proteins without gaps probably gained amino acids via
\emph{insertions}.

@@

A hidden Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse maps each amino acid of the sequence under test onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967hq}.

@@

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{verbatim}
viterbi s n o
  -- consume an observation AND a node
  | s == mat = let trans t prev = 
                   (score + tProb (n-1) t + (eProb n o), mat:path)
                    where (score, path) = viterbi prev (n - 1) (o - 1)
                in minimum $ [
                    trans m_m mat,
                    trans i_m ins,
                    trans d_m del
                    ]
\end{verbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

>>>>

\begin{abstract}
In computational biology, the languages of choice for solving ad-hoc problems are high-level
``scripting'' languages such as Python or domain-specific statistics languages such as R,
while the language of choice for solving high-performance problems is C++.
This dichotomy creates a difficulty when computational biologists wish to experiment with a
variety of approaches to a problem, only some of which are likely to prove fruitful.
This experience report details our use of Haskell in implementing a high-performance,
computational biology application which is the first implementation of a particular
novel algorithm for protein structure and function prediction. We discuss the highlights
and difficulties of this project, as well as the potential improvements the Haskell
community might make that would ease some of these difficulties.
\end{abstract}

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

@@

Homology-detection software is most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. 
This alignment represents structural
correspondence among individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.

An alignment may contain \emph{gaps} (notated as dashes in
\figref{alignment}) for some proteins.
A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
protein~2 lost its amino acid in position~$j$, or 
other proteins gained an amino acid in position~$j$.
If~column~$j$ contains few gaps, 
it~is considered a \emph{consensus column},
and the few proteins with gaps probably lost amino acids via
\emph{deletions}.\remark{NMD: Missing is the fact that all these models
are directionless; they don't care whether something was gained or lost
over time. Perhaps this doesn't matter.}
If~column~$j$ contains \emph{mostly} gaps, 
it~is considered a \emph{non-consensus column},
and the few proteins without gaps probably gained amino acids via
\emph{insertions}.

@@

A hidden Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse maps each amino acid of the query sequence onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967hq}.

@@

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{verbatim}
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{verbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
